\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-06-26 submission/manuscript.tex
update

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

@@

Machine learning (ML) modeling of the human microbiome is used to
identify the microbial biomarkers and aid in diagnosis of many chronic
diseases such as inflammatory bowel disease, diabetes and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes from bacterial abundances, but reliable ML models are
scarce in part due to the flawed modeling methods. Furthermore, using
black box ML models has hindered the validation of microbial biomarkers.
To overcome these challenges, we benchmarked seven different ML models
that use fecal 16S rRNA sequences to predict the presence/absence of
colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
cases). To show the effect of model selection, we assessed the accuracy,
interpretability, and computational efficiency of the following models:
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest, and extreme gradient boosting (XGBoost). The random
forest model was best at detecting CRC lesions with an AUROC of 0.695
but it was slow to train (101.3 h) and hard to interpret. Despite its
simplicity, L2-regularized logistic regression followed random forest in
predictive performance with an AUROC of 0.680, and it trained the
fastest (12 min). This study showed that we should choose ML models
based on our expectations of performance and interpretability as well as
our computational resources. It also established standards for modeling
pipelines of microbiome-associated ML models.

@@

@@

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably,
flawed ML practices are prevalent such as using ML pipelines where there
is no seperate held-out test dataset to evaluate model performance,
reporting few or only the best outcomes of different randomizations of
cross-validation and showing a disregard for large differences between
cross-validation and testing performances as well as large confidence
intervals of testing performances (4, 13--19). Moreover, there is a lack
of discussion on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models such as random forest,
extreme gradient boosting and neural networks without a discussion on if
and how much model interpretibility is necessary for the study (11,
20--22).

The lack of transparency on modeling methodology and model selection
negatively impact model reproducibility and reliability. We need to
strive toward better machine learning practices by (1) implementing
consistent and reliable machine learning pipelines and (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase a reliable ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. CRC is the third most common form of cancer worldwide in which
the human-associated human microbiome is hypothesized to directly
contribute to its development. We built ML models using fecal 16S rRNA
gene sequences to predict patients with normal colons or patients with
colonic tumors which are called screen relevant neoplasias (SRN). The
study had 261 normal and 229 SRN samples. We established modeling
pipelines for L2-regularized logistic regression, L1 and L2 support
vector machines (SVM) with linear and radial basis function kernels, a
decision tree, random forest and XGBoost. Our ML pipeline utilized
held-out test data to evaluate generalization performance of each ML
model. The median test AUROC varied from 0.601 to 0.695. Random forest
had the highest median AUROC for detecting SRN. Despite its simplicity,
the L2-regularized logistic regression followed random forest in
performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
took the longest (162.843 hours, std ± 3.986). We found that median
cross-validation and testing AUROC varied 0.021, which highlights the
importance of a seperate held-out test set and consistent preprocessing
of the data prior to evaluation. Aside from evaluating generalization
and classification performances for each of these models, this study
established standards for modeling pipelines of microbiome-associated
machine learning models.

@@

For regularized logistic regression and SVM with linear kernel we used
L2 regularization to keep all potentially important features. For
comparison, we also trained an L1 regularized SVM model with linear
kernel. L1-regularization on microbiome data lead to a sparser solution
(i.e., force many coefficients to zero), removing features that could be
correlated with other important features. Finally, to explore the
potential for non-linear relationships among features and the outcome of
interest, we trained tree based models, decision tree, random forest and
XGboost, as well as an SVM with non-linear kernel.

We established a robust and reliable ML pipeline where we train and
validate each of the seven models {[}Figure 1{]}. We randomly split the
data into training/validation and test sets so that the
training/validation set consisted of 80\% of the full dataset while the
test set was composed of the remaining data {[}Figure 1{]}. Since the
cases are not uniformly represented in the data, the initial data-split
was stratified to maintain the overall label distribution in both the
training/validation and test sets. Training/validation set consisted of
393 patients (209 SRN), while the test set was composed of 97 patients
(52 SRN). The training/validation data was used for training purposes
and validation of hyperparameter selection (i.e.~model selection), and
the test set was used for evaluation purposes. Validation of
hyperparameter selection was performed using repeated five-fold
cross-validation on the training/validation set {[}Figure 1{]}. Similar
to the initial data-split, five-fold cross-validation was also
stratified to maintain the overall label distribution on the training
and validation sets. We validated the cross-validation performances of
each hyperparameter setting over the 100 randomizations and selected the
best performing hyper-parameter setting in terms of AUROC to train the
full training/validation dataset {[}Figures S1 and S2{]}. We then used
the held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model prediction performance {[}Figure 1{]}.

\textbf{Discriminative performance of the seven models.}

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using the area under the
receiver operating characteristic curve (AUROC) as the discriminative
performance metric {[}Figure 1{]}. Random forest had significantly
higher test AUROC values than the other models for detecting SRNs when
AUROC values were compared to the other six by Wilcoxon rank sum test (p
= 0.0011). The median AUROC of therandom forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another. They had median AUROC values
of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1 SVM with linear kernel and decision tree had
significantly lower AUROC values than the other ML models with median
AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
2{]}.

\textbf{Generalizability of the seven models.}

We compared the median cross-validation AUROC and median testing AUROC
for each model. This difference should be low to suggest the model is
not overfitting despite the large number of features. The largest
difference between the two was 0.021 in L1 SVM with linear kernel
followed by SVM with radial basis function kernel and decision tree with
a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.

We reported the testing AUROC values over 100 random splits of the full
dataset {[}Figure 1{]}. The testing AUROC values varied 0.23 on average,
depending on the data-split. For instance, the lowest AUROC value of the
random forest model was 0.59 whereas the highest was 0.81.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel using the feature weights and L2 logistic
regression using regression coefficients of the trained models. We
examined the 5 predictive (highest weight) and protective (lowest
weight) features identified by the model. We calculated the mean weights
and coefficients of these features over the 100 data-splits. In the
three linear models, OTUs that had the largest mean weights and drove
the detection of SRNs belong to family \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
OTU00050, OTU00015, OTU00768, OTU00822, OTU00609, OTU01212, OTU00629)
and genera \emph{Gamella} (OTU00426) {[}Figure 3{]}. We explained the
feature importances in non-linear models using permutation importance on
the held-out test data where we permuted groups of correlated features
(see methods). For the tree-based models, permuting
\emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
predictive performances the most. Decision tree, random forest and
XGBoost models' predictive performance dropped from 0.6 base testing
AUROC median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
respectively {[}Figure 4{]}. The negative predictive impact of
\emph{Peptostreptococcus} in the decision tree model was followed by a
\emph{Lachnospiraceae} species (OTU00058) (0.6 base testing AUROC median
to 0.58) {[}Figure 4B{]}. Other OTUs had none to minimal effect on the
predictive performance.

@@

Linear models trained faster than non-linear models. L2 logistic
Rregression and L1 and L2 SVM with linear kernel had training times of
0.2 hours, (std ± 0.02), 0.2 hours, (std ± 0.03), and 0.2 hours, (std ±
0.03), respectively. Whereas, SVM with radial basis function kernel,
decision tree, random forest and xgboost had training times of 9.4
hours, (std ± 0.8), 64.7 hours, (std ± 9.9), 101.3 hours, (std ± 10) and
162.8 hours, (std ± 4), respectively {[}Figure 4{]}.

@@

\paragraph{Interpretation of results}\label{interpretation-of-results}

\begin{itemize}
\item
  In this study we established a robust ML pipeline to use 16S rRNA
  sequence counts to predict a binary health outcome. We used a held-out
  test set to illustrate the difference between cross-validation and
  testing AUROC values. When the differences between cross-validation
  and test performance is low, this suggets the models will perform
  similar with similar data. In all seven models, the differences in
  cross-validation and testing AUROC values did not exceed 0.021 which
  suggests that these models will be able to test similar new data.
\item
  We also reported the results of 100 random dataset splits because
  there are differences in AUROC values of each of the seven models
  between different random data-splits. Our results showed that the
  testing AUROC values varied 0.23 on average, depending on the
  data-split. This result shows that you can get ``lucky'' in terms of
  prediction performance depending on the random split. Therefore, it is
  crucial to report all the findings over many random data-splits to get
  reliable performance metrics.
\item
  Our results also showed that we can choose different models for
  different reasons. Getting feature importances as weights and
  regression coefficients is the fastest and easiest interpratation.
  Similarly, linear models require less computational burden, fastest to
  train. Thinking about the applicability, and actionability of a model
  in clinical setting, we might want faster results.
\item
\item
  Suprisingly, l2 logistic regression does really well.
\item
  Computational efficiency and interp of logistic vs random forest
\item
  decision tree and l1 why so low?
\end{itemize}

@@

\begin{itemize}
\item
  What happens with imbalanced data
\item
  What happens with smaller datasets
\item
  Does the code work on any microbiome data?
\item
  Why did we use only one dataset?
\item
  Why not validate on other data instead of held-out testing set?
  Becuase we are comparing several methods on 1 dataset.
\item
  What will the future data look like?
\end{itemize}

@@

\textbf{Data collection and study population.} The data used for this
analysis are stool bacterial abundances and clinical information of the
patients recruited by Great Lakes-New England Early Detection Research
Network study. These data were obtained from Sze et al (23). The stool
samples were provided by recruited adult participants who were
undergoing scheduled screening or surveillance colonoscopy.
Colonoscopies were performed and fecal samples were collected from
participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
labeled by colonoscopy with adequate preparation and tissue
histopathology of all resected lesions. Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as advanced adenoma. Study had 172
patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(24--26).

@@

For L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels we tuned
the \textbf{cost} hyperparameter which determines the regularization
strength where smaller values specify stronger regularization. For SVM
with radial basis function kernel we also tuned \textbf{sigma}
hyperparameter which determines the reach of a single training instance
where for a high value of sigma, the SVM decision boundary will be
dependent on the points that are closest to the decision boundary. For
the decision tree model, we tuned the \textbf{depth of the tree} where
deeper the tree, the more splits it has. For random forest, we tuned the
\textbf{number of features} to consider when looking for the best tree
split. For xgboost, we tuned for \textbf{learning rate} and the
\textbf{fraction of samples} to be used for fitting the individual base
learners. Models were trained using the machine learning wrapper caret
package (v.6.0.81) in R (v.3.5.0).

\textbf{Permutation importance workflow.}

@@

\hypertarget{ref-sze_leveraging_2018}{}
23. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
24. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\hypertarget{ref-westcott_opticlust_2017}{}
25. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\hypertarget{ref-rognes_vsearch_2016}{}
26. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}

@@

Machine learning (ML) modeling of the human microbiome has the potential
to identify the microbial biomarkers and aid in diagnosis of many
chronic diseases such as inflammatory bowel disease, diabetes and
colorectal cancer. Progress has been made towards developing ML models
that predict health outcomes from bacterial abundances, but rigourous ML
models are scarce due to the flawed modeling methods that call into
question the validity of developed ML models. Furthermore, the use of
black box ML models has hindered the validation of microbial biomarkers.
To overcome these challenges, we benchmarked seven different ML models
that use fecal 16S rRNA sequences to predict the presence/absence of
colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
cases). To show the effect of model selection, we assessed the
predictive performance, interpretability, and computational efficiency
of the following models: L2-regularized logistic regression, L1 and L2
support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest, and extreme gradient boosting
(XGBoost). The random forest model was best at detecting CRC lesions
with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to
interpret. Despite its simplicity, L2-regularized logistic regression
followed random forest in predictive performance with an AUROC of 0.680,
and it trained much faster (12 min). This study showed that we should
choose ML models based on our expectations of predictive performance,
interpretability and our computational resources. It also established
standards for modeling pipelines of microbiome-associated ML models.

@@

Prediction of health outcomes using ML is rapidly being adopted by
human-associated microbiome studies. However, the developed ML so far
are overoptimistic in terms of predictive performance and validity.
Without rigorous ML pipelines, we cannot trust ML models. Before we can
speed up progress, we need to slow down, define and start using good ML
practices.

@@

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, we
commonly see ML practices such as using ML pipelines where there is no
seperate held-out test dataset to evaluate model performance or
reporting few or only the best outcomes of cross-validation. Even when
there are seperate testing sets to evaluate model performance, we
observe large differences between cross-validation and testing
performances that indicate overfitting as well as large confidence
intervals for testing performances (4, 13--19). Moreover, there is a
lack of discussion on why a particular ML model is utilized. Recently,
there is a trend towards using more complex ML models such as random
forest, extreme gradient boosting and neural networks without a
discussion on if and how much model interpretibility is necessary for
the study (11, 20--22). Black box machine learning models require
posthoc explanations to determine the importances of features in making
a prediction. These explanations can be misleading and at times
unreliable when making high-stake decisions about someone's health (23).
The lack of transparency on modeling methodology, model interpretation
and model selection negatively impact model reproducibility and
validity. We need to strive toward better machine learning practices by
(1) implementing consistent and rigourous machine learning pipelines and
(2) selecting ML models that reflect the goal of the study as it will
inform our expectations of model accuracy, complexity, interpretibility
and computational efficiency.

To showcase a rigorous ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and the same ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. CRC is a type of cancer which the human-associated human
microbiome is hypothesized to directly contribute to its development
therefore fecal 16S rRNA gene sequences can be used to detect CRC. We
built ML models using fecal 16S rRNA gene sequences to predict healthy
patients versus patients with colorectal lesions that were identified by
colonoscopy as screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate predictive performance and generalizability of each ML
model. We used the area under the receiver operating characteristic
curve (AUROC) as the predictive performance metric. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite its simplicity, the L2-regularized
logistic regression was second best in predictive performance. In terms
of computational efficiency, L1 SVM with linear kernel trained the
fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
(155.104 hours, std ± 0.959). We also found that depending on how the
data is split to create a held-out test set, the AUROC values of a ML
model can vary up to 0.295 which highlighted the importance of
performing many randomized modeling runs. This study established
standards for microbiome-associated ML models and underscored the
importance of model selection.

@@

For regularized logistic regression and SVM with linear kernel we used
L2 regularization to keep all potentially important features. For
comparison, we also trained an L1 regularized SVM model with linear
kernel. L1-regularization on microbiome data lead to a sparser solution
(i.e., force many coefficients to zero). Finally, to explore the
potential for non-linear relationships among features and the outcome of
interest, we trained tree based models, decision tree, random forest and
XGboost, as well as an SVM with non-linear kernel.

We established a robust ML pipeline where we train and validate each of
the seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection (i.e.~model selection), and the test set was
used for evaluation purposes. Validation of hyperparameter selection was
performed using repeated five-fold cross-validation on the
training/validation set {[}Figure 1{]}. Similar to the initial
data-split, five-fold cross-validation was also stratified to maintain
the overall label distribution on the training and validation sets. We
validated the cross-validation performances of each hyperparameter
setting over the 100 randomizations and selected the best performing
hyper-parameter setting in terms of AUROC to train the full
training/validation dataset {[}Figures S1 and S2{]}. We then used the
held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model prediction performance {[}Figure 1{]}.

\textbf{Discriminative performance and generalizability of the seven
models.}

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using AUROC metric {[}Figure
1{]}. Random forest had significantly higher test AUROC values than the
other models for detecting SRNs when AUROC values were compared to the
other six by Wilcoxon rank sum test (p \textless{} 0.01). The median
AUROC of the random forest model was 0.695 (IQR 0.044). L2-regularized
logistic regression, XGBoost, L2-regularized SVM with linear and radial
basis function kernel AUROC values were not significantly different from
one another. They had median AUROC values of 0.68 (IQR 0.055), 0.679
(IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1
SVM with linear kernel and decision tree had significantly lower AUROC
values than the other ML models with median AUROC of 0.65 (IQR 0.066)
and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.

For each model, we compared the median cross-validation AUROC to the
median testing AUROC. The difference between the two should be low to
suggest the model is not overfitting despite the large number of
features. The largest difference between the two was in L1 SVM with
linear kernel, 0.021, followed by SVM with radial basis function kernel
and decision tree with a difference of 0.007 and 0.006, respectively
{[}Figure 2{]}.

We reported the testing AUROC values over 100 random splits of the full
dataset which showed that depending on the data-split, the testing AUROC
value showed great variability {[}Figure 2{]}. The testing AUROC values
within each model varied 0.23 on average across the seven models. For
instance, the lowest AUROC value of the random forest model was 0.59
whereas the highest was 0.81.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel and L2 logistic regression using the feature
weights of the trained models. We ranked the absolute weights of all the
OTUs for each data-split. We calculated the median ranks of these
features over the 100 data-splits. In the three linear models, OTUs that
had the largest median ranks and drove the detection of SRNs belong to
family \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609),
genera \emph{Gamella} (OTU00426) and genera \emph{Peptostreptococcus}
(OTU00367) {[}Figure 3{]}. Some of the OTUs with the highest ranks were
shared among the linear models. We explained the feature importances in
non-linear models using permutation importance on the held-out test data
where we randomly permuted non-correlated features individually and
groups of correlated features together (see methods) to calculate the
effect of permuted OTUs or group of OTUs on testing AUROC. The top 5
OTUs with the largest negative impact on testing AUROC overlapped in
tree-based models {[}Figure 4{]}. Specifically, permuting
\emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
predictive performances the most for all tree-based methods {[}Figure
S3{]}. Decision tree, random forest and XGBoost models' predictive
performance dropped from 0.6 base testing AUROC median to 0.52, from
0.69 to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.
Decision tree was more susceptible to The negative predictive impact of
\emph{Peptostreptococcus} in the decision tree model was followed by a
\emph{Lachnospiraceae} species (OTU00058) (0.6 base testing AUROC median
to 0.58) {[}Figure 4B{]}.

@@

As the complexity of a ML model increased {[}Table S1{]}, its training
times increased as well {[}Figure 5{]}. Additionally, the more
hyperparameters a model had {[}Table S1{]}, the longer it took to train
{[}Figure 5{]}. Linear models trained faster than non-linear models. L2
logistic Rregression and L1 and L2 SVM with linear kernel had training
times of 0.2 hours, (std ± 0.03), 0.2 hours, (std ± 0.02), and 0.2
hours, (std ± 0.02), respectively. Whereas, SVM with radial basis
function kernel, decision tree, random forest and xgboost had training
times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ± 8.8), 83.2 hours,
(std ± 11.3) and 155.1 hours, (std ± 1), respectively {[}Figure 5{]}.

@@

In this study we established a robust ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We used a held-out
test set to illustrate the difference between cross-validation and
testing AUROC values. When the differences between cross-validation and
test performance is low, this suggets the models will perform similar
with similar data. In all seven models, the differences in
cross-validation and testing AUROC values did not exceed 0.021 which
suggests that these models will be able to test similar new data.

In our ML pipeline we performed the 80\%-20\% random datasplit 100
times. The repetition of the initial data-split to create a held-out
test set is a crucial step in the ML pipeline to develop generalizable
ML models and to report reliable performance metrics for a ML model.
Depending on the data-split, there is the chance of being overoptimistic
about the predictive performance of a model. In our study, we showed
that there are differences in AUROC values between different random
data-splits in each of the models we tested. Our results showed that the
testing AUROC values varied 0.23 on average, depending on the
data-split.

Our results also showed that we should choose to use ML models based on
the goal of the study and our expectations of predictive performance,
interpretibility and computational burden. In terms of predictive
performance, all models except a decision tree and L1-regularized SVM
with linear kernel behaved similarly. Random forest model had testing
AUROC values statistically significantly larger than others. However the
second best model was L2-regularized logistic regression following
random forest with a median AUROC difference of only 0.015. The decision
tree model performed poorly because it had no regularization.
L1-regularized SVM with linear kernel also performed poorly possibly
because it was more sensitive to hyper-parameter tuning. In terms of
interpretation, random forest is a more complex ML model and it can only
be explained using methods such as permutation importance. On the other
hand, L2-regularized logistic regression is easy to interpret
(i.e.~regression coefficients of the trained model). If the goal of the
study is to learn the ecology behind a disease and to identify
biomarkers, then easily interpretable models could be a better fit. It
is also important to consider the computational burden of developing ML
models. Random forest model was slow to train with upto \texttt{r} hours
whereas L2-regularized logistic regression trained in \texttt{r}
minutes.

Our study underscored the importance of performing a full grid search
for hyperparameter settings when building a ML model. We first performed
a granular grid search and then a full grid search on the limited range
of possible hyperparameter settings. In the example of L1-regularized
SVM with linear kernel, the model showed variability between different
regularization coefficients (C).

@@

\begin{itemize}
\item
  What happens with imbalanced data The metric you use as the
  discriminative performance becomes important. Don't use accuracy.
  AUROC is good and random at 0.5 always.
\item
  What happens with smaller datasets Generalization performance depends
  on size. The more complex the model, more data you will need. Use
  linear models if you have little data.
\item
  Does the code work on any microbiome data? So if yu have a new sample,
  do you have the same OTUs? NO
\item
  Why did we use only one dataset?
\item
  Why not validate on other data instead of held-out testing set?
  Becuase we are comparing several methods on 1 dataset.
\item
  What will the future data look like?
\end{itemize}

@@

\textbf{Data collection and study population.} The data used for this
analysis are stool bacterial abundances and clinical information of the
patients recruited by Great Lakes-New England Early Detection Research
Network study. These data were obtained from Sze et al (24). The stool
samples were provided by recruited adult participants who were
undergoing scheduled screening or surveillance colonoscopy.
Colonoscopies were performed and fecal samples were collected from
participants in four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
labeled by colonoscopy with adequate preparation and tissue
histopathology of all resected lesions. Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as advanced adenoma. Study had 172
patients with normal colonoscopies, 198 with adenomas and 120 with
carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Stool provided by the patients was used for 16S rRNA gene
sequencing to measure bacterial population abundances. The bacterial
abundance data was generated by Sze et al, by processing 16S rRNA
sequences in Mothur (v1.39.3) using the default quality filtering
methods, identifying and removing chimeric sequences using VSEARCH and
assigning to OTUs at 97\% similarity using the OptiClust algorithm
(25--27).

@@

Models were trained using the machine learning wrapper caret package
(v.6.0.81) in R (v.3.5.0). Within the caret package, we have made
modifications to L2-regularized SVM with linear kernel function
\textbf{svmLinear3} and developed a L1-regularized SVM with linear
kernel function \textbf{svmLinear4} to calculate decision values instead
of predicted probabilities. These changes are available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXXX_2019/}.

For L2-regularized logistic regression, L1 and L2 support vector
machines (SVM) with linear and radial basis function kernels we tuned
the \textbf{cost} hyperparameter which determines the regularization
strength where smaller values specify stronger regularization. For SVM
with radial basis function kernel we also tuned \textbf{sigma}
hyperparameter which determines the reach of a single training instance
where for a high value of sigma, the SVM decision boundary will be
dependent on the points that are closest to the decision boundary. For
the decision tree model, we tuned the \textbf{depth of the tree} where
deeper the tree, the more splits it has. For random forest, we tuned the
\textbf{number of features} to consider when looking for the best tree
split. For XGBoost, we tuned for \textbf{learning rate} and the
\textbf{fraction of samples} to be used for fitting the individual base
learners.For hyperparameter selection, we started with a granular grid
search. Then we narrowed and fine-tuned the range of each
hyperparameter. The range of the grid depends on the ML task and ML
model. A full grid search needs to be performed to avoid variability in
testing performance. We can use hyper-band to help us with our
hyperparameter selection (28).

The computational burden during model training due to model complexity
was reduced by parallelizing segments of the ML pipeline. In this study
we have parallelized each data-split which allowed 100 data-splits to be
processed through the ML pipeline at the same time for each model. We
can further parallelize the cross-validation step for each
hyper-parameter setting.

\textbf{Permutation importance workflow.} We created a Spearman's
rank-order correlation matrix, corrected for multiple pairwise
comparisons. We then defined correlated OTUs as having perfect
correlation (correlation coef=1 and p\textless{}0.01).Other OTUs were
permuted individually to get permutation importance but the correlated
ones are grouped together and permuted at the same time. The reported
OTUs that have impact on testing AUROC are reported in Figures xxx. If
we want we can decrease the correlation coefficient to consider OTUs
that are correlated with ecological consequences but this was out of the
scope of this study but will be followed up in further analyses.

@@

\hypertarget{ref-rudin_please_2018}{}
23. \textbf{Rudin C}. 2018. Please stop explaining black box models for
high stakes decisions. arXiv:181110154 {[}cs, stat{]}.

\hypertarget{ref-sze_leveraging_2018}{}
24. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-schloss_introducing_2009}{}
25. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\hypertarget{ref-westcott_opticlust_2017}{}
26. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\hypertarget{ref-rognes_vsearch_2016}{}
27. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

\hypertarget{ref-li_hyperband:_2016}{}
28. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
bandit-based approach to hyperparameter optimization. arXiv:160306560
{[}cs, stat{]}.
